Monte Carlo Synthetic Acceleration Solutions for the $SP_N$ Equations

### 2013-02-28 doc/spn/spn.tex
adding intro to spn doc

\date{\today}
\title{Monte Carlo Solutions for the $SP_N$ Equations}
\begin{document}
\maketitle

%%---------------------------------------------------------------------------%%
\section{Introduction}
\label{sec:introduction}

>>>>

\date{\today}
\title{Monte Carlo Synthetic Acceleration Solutions for the $SP_N$ Equations}
\begin{document}
\maketitle

%%---------------------------------------------------------------------------%%
\section{Introduction}
\label{sec:introduction}
The neutron tranpsort problem is complicated. Solutions cover a large
phase space and the problems of interest are often geometry complex,
very large, or both requiring tremendous computational resources to
generate an adequate solution. Modern deterministic methods for large
scale problems are commonly variants on the discrete ordinates ($S_N$)
method \cite{evans_denovo:_2010}. For full reactor core neutronics
simulations, the $S_N$ method requires potentially trillions of
unknown angular flux moments to be computed to achieve good accuracy
for the repsonses to be computed. Other forms of the transport
problem, including the $P_N$ method, take on a simpler form than the
more common $S_N$ methods but lack in accuracy when compared while
still requiring considerable computational resources for solutions in
multiple dimensions.

In the 1960's, Gelbard developed an ad-hoc multidimensional extension
of the simple single dimension planar $P_N$ equations that created a
system of coupled, diffusion-like equations known as the $SP_N$
equations. Up until around the 1990's, the $SP_N$ methods was either
widely unknown, widely unused, or combination of both even though
numerical studies showed promising results with better solutions than
diffusion theory and a significant reduction in computational time
over more accurate methods such as discrete ordinates. Why did this
happen? A significant problem, pointed out by Larsen, was that little
rigor had been applied to the formulation of the $SP_N$ equations
since their derivation through primarily heuristic arguments. Instead,
studies at that time focused on simply comparing the results of the
method to other contemporary transport solution strategies. In
addition, many problems of interest from the literature at the time
were either solved using nodal-type methods for reactor-sized problems
are $S_N$-type methods for benchmark problems with intricate material
configurations and potentially large flux gradients over small spatial
domains.

So why reconsider the $SP_N$ equations? Starting in the 1990's and
primarily due to Larsen and his colleagues, the $SP_N$ equations have
been given a more rigorous treatment with both variational and
asymptotic derivations performed as a means of verification. In
addition, these equations have been more rigorously studied as
solution methods to MOX fuel problems and have been shown to provide
accurate solutions. With this mathematical literature to provide a
solid numerical footing for the method, we look at its application to
today's challenge problems in full reactor core transport. The
reduction in numerical complexity of current full core deterministic
solution methods using the $S_N$ approximation could mean significant
savings in both compute time and memory required. In addition, the
characteristics of the solution to the transport problem for a steady
state reactor core permit diffusion theory to be used; a staple of the
nuclear industry since its inception. Therefore, if diffusion theory
is applicable, then finer grained solutions that capture more of the
physics contained in the transport equation should be possible with
the $SP_N$ method. In doing so, we also expect from the literature to
obtain computed responses on the order of accuracy we would expect
from an appropriately discretized $S_N$ method at a fraction of the
cost.

To further motivate moving in this direction, recent developments in
the Exnihilo neutronics package at Oak Ridge National Laboratory have
permitted generation of the $SP_N$ system of equations for detailed
full core reactor models. By fully forming these equations and
formulating them as a linear algebra problem, we now have access to
all of the modern advancements in computational linear algebra
including Krylov solvers for asymmetric systems and preconditioning
methods such as algebraic multigrid. This leads us to then explore the
applicability of our work in discrete Monte Carlo methods for linear
systems as a possible solution method for the $SP_N$ equations. If
formulated correctly, we hypothesis that signficant improvement in
usage of computational resources may be observed compared to modern
solution techniques such as those suggested.

In this document we derive the multigroup $SP_N$ equations from the
Boltzmann transport equation. We then perform a spectral analysis on
the resulting system of equations to determine its potential
performance with both modern Krylov methods and Monte Carlo methods
for asymmetric systems. Following this, we devise a Monte Carlo
synthetic acceleration method as a proposed solution scheme.
